### 2012-05-10 Rctl/inst/doc/manual.Rnw
[Rctl] Manual updates, add bigdata and ctlscan

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Danny Arends
% The ctl mapping routine
%\VignetteIndexEntry{Manual}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[12pt]{article}

@@

@@

After you have stored the data as plain text open the files in a text editor (NOT WORD, for 
windows Notepad++ is a good option) and check to see if your data is stored correctly, if 
not reformat your data in both the genotype.csv and the phenotype.csv files to match the 
format below. {\color{red}\emph{\bf{NOTE: Individuals must match between the genotype and 
the phenotype file...}}} The phenotypes.csv file containing the genotype matrix is stored 
individuals x genetic marker:

\begin{tabular}{lrrrrr}
      & Marker1 & Marker2 & Marker3 & ... & MarkerN \\
Ind1  & 1 & 1 & 2 & ... & 2 \\
Ind2  & 1 & 1 & 2 & ... & 2 \\
Ind3  & 1 & 1 & 1 & ... & 2 \\
...   & ... & ... & ... & ... & ... \\
IndN  & 1 & 1 & 1 & ... & 2 \\
\end{tabular}

@@

\emph{Adding genetic map information}
This step is optional, but plots and generated networks look much nicer when chromosome locations
of genetic markers are supplied. The 'mapinfo' object is matrix with 3 columns: "Chr" - the 
chromosome number, "cM" - the location of the marker in centiMorgans and the 3rd column "Mbp" - The 
location of the marker in Mega basepairs. The structure of the $mapinfo.csv$ file:

@@

<<echo=TRUE,eval=FALSE>>=
mapinfo <- read.csv("mapinfo.csv",row.names=1,col.names=TRUE)
traits[1:5,1:3] #Show the first 5 marker records
@
After this we can
\bigskip

\textbf{\sffamily CTLscan and CTLscan.network}
\nopagebreak

Explains how to map CTL using the CTLscan and CTLscan.network functions.
<<echo=TRUE,eval=FALSE>>=
require(ctl)        #Loads the R/ctl package
data(ath.metab)
geno <- ath.metab$genotypes #Short name
traits <- ath.metab$phenotypes #Short name

#Scan all phenotypes for CTLs, using the default options
#NOTE: Could take some time
ctls <- CTLscan(geno,traits)
@

\emph{CTLscan options}
The genotype matrix and the $geno.enc$ param, the phenotype matrix, the important $n.perms$ param, 
which should be set to as many as possible, especially when publishing any results, but for a quick 
scan 100 is ussualy sufficient.

When experimental setups are more complex a user can supply QTL results via the $have.qtl$ option.

<<echo=TRUE,eval=FALSE>>=
require(qtl)        #load the R/qtl package
data(multitrait)    #load the dataset
qtls <- scanone(multitrait, pheno.col=1:24)[,-c(1,2)]
ctls <- CTLscan.cross(multitrait, pheno.col=1:2, have.qtl=qtls)
@

, when no qtls are supplied the internal QTL mapping routine is used, which can deal with just a 
single condition per individual. An environmental condition can be specified by the $conditions$ 
param and are also passed to the internal QTL mapping algorithm.

@@

\bigskip
\textbf{\sffamily Advanced plots and CTL networks}
\nopagebreak

@@

Explains how to use the D version to map big datsets

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Danny Arends
% The ctl mapping routine
%\VignetteIndexEntry{Manual}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[10pt]{article}

@@

\newpage

@@

After you have stored the data as plain text open the files in a text editor (NOT WORD, for 
windows Notepad++ is a good option) and check to see if your data is stored correctly, if 
not reformat your data in both the genotype.csv and the phenotype.csv files to match the 
format below. {\color{red} \emph{\bf{NOTE: Individuals must match between the genotype and 
the phenotype file...}}} The phenotypes.csv file containing the genotype matrix is stored 
individuals x genetic marker:

\begin{tabular}{lrrrrr}
      & Marker1 & Marker2 & Marker3 & ... & MarkerN \\
Ind1  & 1 & 1 & 1 & ... & 2 \\
Ind2  & 1 & 2 & 2 & ... & 1 \\
Ind3  & 2 & 2 & 1 & ... & 1 \\
...   & ... & ... & ... & ... & ... \\
IndN  & 1 & 1 & 1 & ... & 2 \\
\end{tabular}

@@

\emph{Adding genetic map information}\\
This step is optional, but plots and generated networks look much nicer when chromosome locations
of genetic markers are supplied. The 'mapinfo' object is matrix with 3 columns: "Chr" - the 
chromosome number, "cM" - the location of the marker in centiMorgans and the 3rd column "Mbp" - The 
location of the marker in Mega basepairs. The structure of the $mapinfo.csv$ file:

@@

<<echo=TRUE,eval=FALSE>>=
mapinfo <- read.csv("mapinfo.csv",row.names=1,col.names=TRUE)
traits[1:5,1:3]     #Show the first 5 marker records
@
Finally... after all this we can start scanning QTL and CTL...
\bigskip

\textbf{\sffamily Scanning for CTL / $CTLscan$}
\nopagebreak

We start explaining how to map CTL using the basic options of the $CTLscan$ function. This function 
scan CTL and produces an output "CTLobject".
<<echo=TRUE,eval=FALSE>>=
require(ctl)                     #Loads the R/ctl package
data(ath.metab)                  #Loads the example dataset
geno <- ath.metab$genotypes      #Short name
traits <- ath.metab$phenotypes   #Short name

#Scan all phenotypes for CTLs, using the default options (Could take some time)
ctls <- CTLscan(geno,traits)
@
If you have loaded an R/qtl "cross" object, use the $CTLscan.cross$ interface, this will automagically 
extract the cross, check if the type of "cross" is supported, and deal with the $geno.enc$ parameter:
<<echo=TRUE,eval=FALSE>>=
library(qtl)
data(multitrait)
ctls <- CTLscan.cross(multitrait)
@

Now you might feel tempted to just dive into the results, But... first lets take a look at some 
options that might apply to your experimental setup like treatments applied or growth conditions, 
and which (if set incorrectly) could seriously reduce mapping power. Also you might want to add 
your own/previous QTL results.

\emph{CTLscan options}\\
The genotype matrix and the $geno.enc$ parameter, the phenotype matrix, the important $n.perms$ 
parameter, which should be set to as many as possible, especially when publishing any results, 
but for a quick scan 100 is ussualy sufficient.

When experimental setups are more complex a user can supply QTL results via the $have.qtl$ option.
An example is shown below in which we use QTL results obtained by using single marker mapping via 
$scanone$ and the new Multiple QTL mapping routine $mqmscan$ in R/qtl:

<<echo=TRUE,eval=FALSE>>=
require(qtl)        #load the R/qtl package
data(multitrait)    #load the dataset
oneqtls   <- scanone(multitrait, pheno.col=1:24)[,-c(1,2)]
ctls_one  <- CTLscan.cross(multitrait, pheno.col=1:2, have.qtl=qtls)
@

And a more advanced example where the MQM scan uses 30 genetic markers as cofactors to first build 
up a genetic model underlying the QTL and locus to locus interactions:

<<echo=TRUE,eval=FALSE>>=
cofactors <- mqmautocofactors(multitrait, 30)
mqm_qtls  <- mqmscanone(multitrait, cofactors=cofactors, pheno.col=1:24)[,-c(1,2)]
ctls_mqm  <- CTLscan.cross(multitrait, pheno.col=1:2, have.qtl=qtls)
@

CTLs can be condition dependant just like QTL, so if one has multiple conditions or treatments. we 
need to acomodate that in the underlying models. If no QTLs are supplied the internal QTL mapping 
routine is used, which can deal with just a single condition per individual. An environmental 
condition can be specified by the $conditions$ parameter:

@@

\bigskip
\textbf{\sffamily RESULTS - Plots and CTL networks}
\nopagebreak

@@

Big datsets require special care, R isn't the best platform to deal with large scale genomics data 
from micro- or tilling arrays (Just to mention RNAseq and full DNA strand sequencing). Thus we also
provide a commandline version of out CTL mapping tool: mapctl.exe, for more information on all the 
command line parameters can be found at http://www.mapctl.org/mapctl.html
The mapctl commandline tool uses the same input files as the R version, and supports qtab the new 
genotype and phenotype encoding scheme used by qtlHD.
It provides only scanning and permutation features, the output is analyzed in R as outlined above. 
To load the results from a mapctl scan into R by specifying the input files and the output directory 
created by the tool using:

<<echo=TRUE,eval=FALSE>>=
ctls <- read.dctl("genotypes.csv","phenotypes.csv",results="<PATH>/output/")
@
